A Data-Driven Framework for Neural Field Modelling

### 2010-11-14 ltx/first_order_framework/LaTeX_Manuscript.tex
small general edits to paper, improved some of the results section

An estimation framework for neural mass models known as dynamical causal modelling (DCM) \cite{David2003,David2006} has been proposed for studying evoked potential dynamics. Via a Bayesian inference scheme, DCM estimates the long range (cortico-cortical) connectivity structure between the specific isolated brain regions that best explains a given data set using the Jansen and Ritt equations~\cite{Jansen1995}. Another recent approach estimates the parameters of a neural field model using an unscented Kalman filter based on modified WC neural field equations~\cite{schiff2008kalman}. This work takes a system theoretic approach to the neural estimation problem, successfully demonstrating that it is possible to perform state estimation on spatiotemporal neural fields. This marks the first step in what has the potential to revolutionise the treatment of many neurological diseases where therapeutic electrical stimulation is viable.
\dean{and here}

We present an extension to the work of Schiff and Sauer~\cite{schiff2008kalman} by establishing a framework for estimating the state of the WC equations for larger scale (more space) systems via a systematic model reduction procedure. In addition, a method is presented for estimating the connectivity structure and the synaptic dynamics. Until now, model-based estimation of local intracortical connectivity has not been reported in the literature (to the best of the authors' knowledge). Our study extends recent work which shows that it is possible to estimate local coupling of spatiotemporal systems using techniques from control systems theory and machine learning~\cite{Dewar2009}. The key development of this previous work was to represent a spatiotemporal system as a standard state-space model, with the number of states independent of the number of observations (recording electrodes in this case). In addition, the appropriate model selection tools have been developed~\cite{Scerri2009} allowing for the application of the technique to neural fields. This paper extends the linear framework of \cite{Dewar2009,Scerri2009} to the nonlinear case required by the WC model.

@@

\subsection{Neural Field Model}\label{NeuralModelSection} 
Neural field models relate mean firing rates of pre-synaptic neural populations to mean post-synaptic membrane potentials. They are popular as they are parsimonious yet have a strong link with the underlying physiology. Each neural population represents a functional cortical processing unit, such as a column. The columnar organisation of the cortex is continuous, where pyramidal cells are members of many columns. In general, cortical structure can be modelled in a physiologically plausible manner as being locally homogeneous (in short range intracortical connectivity) and heterogeneous (in long range cortico-cortical and corticothalamic connectivity)~\cite{Jirsa2009,Qubbaj2007}. In certain regions of the cortex, each column is thought to be connected locally via symmetric short range local excitation with surround inhibition \cite{Braitenberg1998}. For example, this structural organisation is most studied in the visual system, where the surrounding inhibition effectively tunes a cortical column to a particular receptive visual field~\cite{Sullivan2006}. Neural field models are descriptive of a range of neurodynamics of the cortex such as evoked potentials, visual hallucinations and epileptic behaviour~\cite{David2003,Bressloff2001,Breakspear2006}. Field models are also capable of generating complex spatial patterns of activity such as Turing patterns, spirals and travelling oscillations~\cite{Amari1977,Coombes2005,Coombes2007}.

@@

\subsection{State and Parameter Estimates} 
\label{sec:state_and_param_results}
Experiments were performed to demonstrate the performance of the state and parameter estimation algorithm. The first experiment evaluates the estimation performance using a Monte Carlo approach for a single set of parameters. The second experiment evaluates the performance with a selection of different connectivity kernel parameters. The third experiment shows the dependance of the estimation accuracy on the parameters of the sigmoidal nonlinearity. The final experiment shows the effect of the width of the sensor kernel on the estimation performance.

@@

Histograms of the parameter estimates obtained over the 150 realisations are shown in Figure~\ref{fig:Figure7}. The actual parameter values, the mean estimates, corresponding standard deviations and biases are given with the results from the next Section in Table~\ref{tab:Parameters estimates}. The biases show the percentage differences between the actual values and the mean parameter estimates. The mean estimates are in good agreement with the actual values of the connectivity kernel parameters, however the time constant parameter, $\xi$, is outside the estimate distribution (with bias of 2.78\%).

@@

To study the robustness of the estimation algorithm, the performance was studied for various connectivity kernel parameters. The reconstructed connectivity kernels from the estimated parameters are shown in Figure~\ref{fig:Figure8}. For Parts a), b) and c) of the figure a Mexican hat connectivity kernel was used. For each subset of parameters the mean of the reconstructed kernel over 50 realisations is shown as a dotted line and the 95~\% confidence interval is indicated by the grey shaded region. In each case the actual kernel was within the 95~\% confidence of kernel estimate. The statistics for the kernel parameters estimates for each of the parameter sets are shown in Table~\ref{tab:Parameters estimates}.

For each of the Mexican hat connectivities, the support of the kernel was assumed to be known. Parts d), e) and f) of Figure~\ref{fig:Figure8} show results when the kernel support in the estimator differed from the actual data. In these cases the data was generated using a single positive (excitatory) Gaussian kernel, as seen by the solid line in the figure. The estimation algorithm was constructed allowing for the Mexican hat connectivity of the previous examples. Each plot shows that the actual kernel lies within the confidence intervals of the reconstructed estimates, for the majority of the kernel space until the amplitude is relatively low. Table~\ref{tab:GaussianKernelResults} shows the statistics of the parameter estimates. Importantly, although the reconstructed kernel estimate is in good accord with the actual kernel, the individual kernel basis function weights are not in agreement with the actual parameters. This result demonstrates that unless the support of the kernel basis functions are precisely known, the exact meaning of the parameters in isolation is not clear and the net reconstructed kernel should be considered in a practical setting.

@@

Figure~\ref{fig:Figure11} demonstrates the affect of the sigmoid parameters on the estimation accuracy. The RMSE of the reconstructed connectivity kernel was calculated for various parameters of the sigmoid by varying the slope, $\varsigma$, and the threshold, $v_0$. The kernel and time constant parameters used in this experiment were the one that gave the best results from the previous experiment. 
% The size of the squares in the figure show the magnitude of the log of the error. The squares are black if the log of the error is negative (the RMSE is $<1$) and the squares are white if the log of the error is positive (the RMSE is $>1$). The largest black squares indicate the best performance and the largest white squares indicate the worst performance for the parameter combinations (\dean{need some thing more quantitative}). 
From the figure it can be seen that the estimation accuracy depends on the characteristics of the sigmoid function. When the threshold is negative or the slope is very gentle, all of the activity is being passed through the sigmoid function. However, as the slope is increased or the threshold is increased, more information is being cut off or attenuated by the activation function. This is illustrated in Figure~\ref{fig:sigmoidfunctions}, where the sigmoids from the bottom left corner through to near top right corner of Figure~\ref{fig:Figure11} are plotted. This appears to have a detrimental effect on the estimation accuracy.

@@

\tiny{\begin{tabular}{lcccccccccccc}
Experiment&\multicolumn{4}{c}{I}&\multicolumn{4}{c}{II}&\multicolumn{4}{c}{III}\\
 Parameters&$\theta_0$&$\theta_1$&$\theta_2$&$\xi$&$\theta_0$&$\theta_1$&$\theta_2$&$\xi$&$\theta_0$&$\theta_1$&$\theta_2$&$\xi$\\
\hline\hline
Actual Value&$90$&$-80$&$5$&$0.9$&$100$&$-80$&$5$&$0.9$&$110$&$-80$&$5$&$0.9$\\
Mean Estimate&$103.86$&$-89.05$&$5.23$&$0.923$&$101.75$&$-81.00$&$4.76$&$0.924$&$98.90 $&$-73.13$&$4.37$&$0.925$\\
Std. Dev.&$23.14$&$16.57$&$0.74$&$0.003$&$21.30$&$14.82$&$0.65$&$0.003$&$16.48$& $11.72$&$0.54$&$0.003$\\
Bias (\%)& $15.4$&$11.31$&$4.6$&$2.56$&$1.75$&$1.25$&$4.8$&$2.67$&$10.1$&$ 8.58$&$12.6$&$2.78$  \\
\end{tabular}}

>>>>

An estimation framework for neural mass models known as dynamical causal modelling (DCM) \cite{David2003,David2006} has been proposed for studying evoked potential dynamics. Via a Bayesian inference scheme, DCM estimates the long range (cortico-cortical) connectivity structure between the specific isolated brain regions that best explains a given data set using the Jansen and Ritt equations~\cite{Jansen1995}. Another recent approach estimates the parameters of a neural field model using an unscented Kalman filter based on modified WC neural field equations~\cite{schiff2008kalman}. This work takes a system theoretic approach to the neural estimation problem, successfully demonstrating that it is possible to perform state estimation and control on spatiotemporal neural fields. This marks the first step in what has the potential to revolutionise the treatment of many neurological diseases where therapeutic electrical stimulation is viable.
\dean{and here}

We present an extension to the work of Schiff and Sauer~\cite{schiff2008kalman} by establishing a framework for estimating the state of the WC equations for larger scale (more space) systems via a systematic model reduction procedure. In addition, a method is presented for estimating the connectivity structure and the synaptic dynamics. Until now, model-based estimation of local intracortical connectivity has not been reported in the literature (to the best of the authors' knowledge). Our study extends recent work which shows that it is possible to estimate local coupling of spatiotemporal systems using techniques from control systems theory and machine learning~\cite{Dewar2009}. The key development of this previous work was to represent a spatiotemporal system as a standard state-space model, with the number of states independent of the number of observations (recording electrodes in this case). In addition, the appropriate model selection tools have been developed~\cite{Scerri2009} allowing for the application of the technique to neural fields. This paper extends the linear framework of~\cite{Dewar2009,Scerri2009} to the nonlinear case required by the WC model.

@@

\subsection{Neural Field Model}\label{NeuralModelSection} 
Neural field models relate mean firing rates of pre-synaptic neural populations to mean post-synaptic membrane potentials. They are popular as they are parsimonious yet have a strong link with the underlying physiology. Each neural population represents a functional cortical processing unit, such as a column. The columnar organisation of the cortex is continuous, where pyramidal cells are members of many columns. In general, cortical structure can be modelled in a physiologically plausible manner as being locally homogeneous (in short range intracortical connectivity) and heterogeneous (in long range cortico-cortical and corticothalamic connectivity)~\cite{Jirsa2009,Qubbaj2007}. In certain regions of the cortex, each column is thought to be connected locally via symmetric short range local excitation with surround inhibition~\cite{Braitenberg1998}. For example, this structural organisation is most studied in the visual system, where the surrounding inhibition effectively tunes a cortical column to a particular receptive visual field~\cite{Sullivan2006}. Neural field models are descriptive of a range of neurodynamics of the cortex such as evoked potentials, visual hallucinations and epileptic behaviour~\cite{David2003,Bressloff2001,Breakspear2006}. Field models are also capable of generating complex spatial patterns of activity such as Turing patterns, spirals and travelling oscillations~\cite{Amari1977,Coombes2005,Coombes2007}.

@@

\subsection{State and Parameter Estimates} 
\label{sec:state_and_param_results}
Experiments were performed to demonstrate the performance of the state and parameter estimation algorithm. The first experiment evaluates the estimation performance using a Monte Carlo approach for a single set of parameters. Using this experiment we demonstrate convergence of the algorithm, show the distribution of estimated parameters and the accuracy of the state estimates for a single set of parameters. The second experiment evaluates the performance with a selection of different connectivity kernel parameters. In addition, results are shown where the support of the connectivity kernels differs between actual data and the estimator. The third experiment shows the how the estimation accuracy varies depending on the parameters of the sigmoidal nonlinearity. The final experiment shows the effect of the width of the sensor kernel on the estimation performance.

@@

Histograms of the parameter estimates obtained over the 150 realisations are shown in Figure~\ref{fig:Figure7}. The actual parameter values, the mean estimates, corresponding standard deviations and biases are given with the results from the next Section in Table~\ref{tab:MexicanHatKernelResults}. The biases show the percentage differences between the actual values and the mean parameter estimates. The mean estimates are in good agreement with the actual values of the connectivity kernel parameters, however the time constant parameter, $\xi$, is outside the estimate distribution (with bias of 2.78\%), with a small variance in parameter estimates.

@@

To study the robustness of the estimation algorithm, the performance was studied for various connectivity kernel parameters. The reconstructed connectivity kernels from the estimated parameters are shown in Figure~\ref{fig:Figure8}. For Parts a), b) and c) of the figure a Mexican hat connectivity kernel was used. For each subset of parameters the mean of the reconstructed kernel over 50 realisations is shown as a dotted line and the 95~\% confidence interval is indicated by the grey shaded region. In each case the actual kernel was within the 95~\% confidence of kernel estimate. The statistics for the kernel parameters estimates for each of the parameter sets are shown in Table~\ref{tab:MexicanHatKernelResults}.

For each of the Mexican hat connectivities, the support of the kernel was assumed to be known. However, in Parts d), e) and f) of Figure~\ref{fig:Figure8} the kernel support in the estimator differed from the actual data. In these cases the data was generated using a single positive (excitatory) Gaussian kernel, as seen by the solid line in each of the subplots of the figure. The estimation algorithm was constructed allowing for the Mexican hat connectivity of the previous examples. Each plot shows that the actual kernel lies within the confidence intervals of the reconstructed estimates, for the majority of the kernel space until the amplitude is relatively low. Table~\ref{tab:GaussianKernelResults} shows the statistics of the parameter estimates. Importantly, although the reconstructed kernel estimate is in good accord with the actual kernel, the individual kernel basis function weights are not in agreement with the actual parameters. This result demonstrates that unless the support of the kernel basis functions are precisely known, the exact meaning of the parameters in isolation is not clear and the net reconstructed kernel should be considered in a practical setting.

@@

Figure~\ref{fig:Figure11} demonstrates the affect of the sigmoid parameters on the estimation accuracy. The RMSE of the reconstructed connectivity kernel was calculated for various parameters of the sigmoid by varying the slope, $\varsigma$, and the threshold, $v_0$. The other model parameters remained fixed and where set according the Table~\ref{tab:Parameters}.

% The size of the squares in the figure show the magnitude of the log of the error. The squares are black if the log of the error is negative (the RMSE is $<1$) and the squares are white if the log of the error is positive (the RMSE is $>1$). The largest black squares indicate the best performance and the largest white squares indicate the worst performance for the parameter combinations (\dean{need some thing more quantitative}). 
From the figure it can be seen that the estimation accuracy is inter-related to the properties of the sigmoid function. When the threshold is negative or the slope is very gentle, all of the activity is being passed through the sigmoid function. However, as the slope is increased or the threshold is increased, more information is being cut off or attenuated by the activation function. This is illustrated in Figure~\ref{fig:sigmoidfunctions}, where the sigmoids from the bottom left corner through to near top right corner of Figure~\ref{fig:Figure11} are plotted. This appears to have a detrimental effect on the estimation accuracy.

@@

\tiny{\begin{tabular}{lcccccccccccc}
Experiment&\multicolumn{4}{c}{I}&\multicolumn{4}{c}{II}&\multicolumn{4}{c}{III}\\
 Parameters&$\theta_0$&$\theta_1$&$\theta_2$&$\xi$&$\theta_0$&$\theta_1$&$\theta_2$&$\xi$&$\theta_0$&$\theta_1$&$\theta_2$&$\xi$\\
\hline\hline
Actual Value&$90$&$-80$&$5$&$0.9$&$100$&$-80$&$5$&$0.9$&$110$&$-80$&$5$&$0.9$\\
Mean Estimate&$103.86$&$-89.05$&$5.23$&$0.923$&$101.75$&$-81.00$&$4.76$&$0.924$&$98.90 $&$-73.13$&$4.37$&$0.925$\\
Std. Dev.&$23.14$&$16.57$&$0.74$&$0.003$&$21.30$&$14.82$&$0.65$&$0.003$&$16.48$& $11.72$&$0.54$&$0.003$\\
Bias (\%)& $15.4$&$11.31$&$4.6$&$2.56$&$1.75$&$1.25$&$4.8$&$2.67$&$10.1$&$ 8.58$&$12.6$&$2.78$  \\
\end{tabular}\label{tab:MexicanHatKernelResults}}
